---
title: |
       <span style="color:white">Descriptive Statistics</span>

author: "Yebelay Berehan <br><span style = 'font-size: 50%;'>Biostatistician </span>"
institute: |
           [<span style = 'font-size: 80%;'> <span style="color:cyan">yebelay.ma\@gmail.com</span></span>](yebelay.ma\@gmail.com)
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: [xaringan-themer.css,  rladies, rladies-fonts]
    nature:
     # highlightStyle: googlecode
      highlightLines: true
      #highlightLanguage: ["r"]
      ratio: "14:9"
     # highlightLines: true
      highlightSpans: true
      highlightStyle: github
      countIncrementalSlides: false
      titleSlideClass: ["center","middle"]
  includes:
    in_header: columns.tex
---
```{r setup, include=FALSE, purl=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "##")
library(datasets)
library(tidyverse)
library(ggplot2)
library(dplyr)
```

```{r xaringan-logo, echo=FALSE}
xaringanExtra::use_logo(image_url = "logo1.png", 
  exclude_class = c("inverse", "hide_logo"),  width = "90px",height = "85px")
```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#562457", 
  secondary_color = "#88398A",
  #text_font_google = google_font("Roboto"),
  text_font_size = "24px",
  background_color = "#FFFFFF",
  #code_font_family = "Ligature Code",
  code_font_size = "0.8em", 
  footnote_color = "blue",
  footnote_position_bottom = "0.1em",
)
```

## Descriptive statistics in R

- Introduction
- Descriptive statistics
   - Min,max,Range
   - Mean,Median, Mode,Quantiles, IQR
   - Sd and variance
   - Summary
  - Contingency table
- Advanced descriptive statistics
- Graphical procedures
  - Barplot, Boxplot, Pie chart
  - Histogram, Scatterplot, Line plot, QQ-plot, Density plot

---

# Introduction

- How to compute the main descriptive statistics in R and how to present them graphically. 

-	descriptive statistics is a branch of statistics aiming at summarizing, describing and presenting a series of values or a dataset. 

-	Descriptive statistics is often the first step and an important part in any statistical analysis. 

- It allows to check the quality of the data and it helps to “understand” the data by having a clear overview of it. 

-	If well presented, descriptive statistics is already a good starting point for further analyses. There exists many measures to summarize a dataset. 

---

They are divided into two types:

-	**location measures** and	**dispersion measures**

-	Location measures give an understanding about the central tendency of the data, whereas dispersion measures give an understanding about the spread of the data. 

- We use the dataset iris and we need to load it by running iris:
```{r}
dat <- iris # load the iris dataset and renamed it dat
```

- Below a preview of this dataset and its structure:
```{r, eval=F}
head(dat) # first 6 observations
str(dat) # structure of dataset
```

-	The dataset contains 150 observations and 5 variables. Length and width of the sepal and petal are numeric variables and the species is a factor with 3 levels. 
---

#### Minimum and maximum

- Minimum and maximum can be found thanks to the min() and max() functions:

```{r, eval=FALSE}
min(dat$Sepal.Length)
max(dat$Sepal.Length)
```

- Alternatively the range() function:
```{r}
rng <- range(dat$Sepal.Length)
```

- gives you the minimum and maximum directly. 

```{r, eval=FALSE}
rng[1] # rng = name of the object specified above
# and the maximum with:
rng[2]
```
---

#### Range

- The range can then be easily computed, as you have guessed, by subtracting the minimum from the maximum:

```{r, eval=FALSE}
max(dat$Sepal.Length) - min(dat$Sepal.Length)
```

- We can create our own function to compute the range:
```{r, eval=FALSE}
range2 <- function(x) {
  range <- max(x) - min(x)
  return(range)
}
range2(dat$Sepal.Length)
```

- which is equivalent than max−minmax−min presented above.

---

#### Mean

- The mean can be computed with the mean() function:
```{r, eval=FALSE}
mean(dat$Sepal.Length)
```

Tips:	if there is missing value in your dataset, use `mean(dat$Sepal.Length, na.rm = TRUE)`.

#### Median

- The median can be computed thanks to the median() function:
```{r, eval=FALSE}
median(dat$Sepal.Length) # or with the quantile() function:
quantile(dat$Sepal.Length, 0.5)
```

- since the quantile of order 0.5 (q0.5) corresponds to the median.
- by setting the second argument to **0.25 or 0.75**, to compute  first and third quartile. 

---

#### Mode

- To my knowledge there is no function to find the mode of a variable. 
- However, we can easily find it thanks to the functions `table()` and `sort()`:

```{r, eval=FALSE}
tab <- table(dat$Sepal.Length) # number of occurrences for each unique value
sort(tab, decreasing = TRUE) # sort highest to lowest
```
- `table()` gives the number of occurrences for each unique value, then `sort()` with the argument `decreasing = TRUE` displays the number of occurrences from highest to lowest. 

- The mode of the variable Sepal.Length is thus 5. 

- This code to find the mode can also be applied to qualitative variables such as Species:
```{r, eval=FALSE}
sort(table(dat$Species), decreasing = TRUE) # or:
summary(dat$Species)
```

---

#### Interquartile range

The interquartile range can be computed with the IQR() function:
```{r, eval=FALSE}
IQR(dat$Sepal.Length)# or using quantile() 
quantile(dat$Sepal.Length, 0.75) - quantile(dat$Sepal.Length, 0.25)
```

#### Standard deviation and variance

- The standard deviation and the variance is computed with the sd() and var() functions:
```{r, eval=FALSE}
sd(dat$Sepal.Length) # standard deviation
var(dat$Sepal.Length) # variance
```

Tip: to compute the standard deviation of multiple variables at the same time, use `lapply()`
```{r, eval=FALSE}
lapply(dat[, 1:4], sd)
```

---

### Summary

- You can compute the minimum, 1st quartile, median, mean, 3rd quartile and the maximum for all numeric variables of a dataset at once using summary():
```{r}
summary(dat, digits = 3)
```

---
- Tip: if you need these descriptive statistics by group use the by() function:
```{r, eval=FALSE}
by(dat, dat$Species, summary)
```

- where the arguments are the name of the dataset, the grouping variable and the summary function. 
- If you need more descriptive statistics, use **stat.desc()** from the package `pastecs`:

```{r, eval=FALSE}
library(pastecs)
stat.desc(dat)
```

- You can have even more statistics (i.e., skewness, kurtosis and normality test) by adding the argument `norm = TRUE` in the previous function. 

- Note that the variable Species is not numeric, so descriptive statistics cannot be computed for this variable and NA are displayed.
---

### Contingency table

- `table()` introduced above can also be used on two qualitative variables to create a contingency table. 

- The dataset iris has only one qualitative variable so we create a new qualitative variable just for this example.

- We create the variable size which corresponds to small if the length of the petal is smaller than the median of all flowers, big otherwise:
```{r, results=FALSE}
dat$size <- ifelse(dat$Sepal.Length < median(dat$Sepal.Length),
  "small", "big")
```
---
- Here is a recap of the occurrences by size:
```{r, results=FALSE}
table(dat$size)
```

- We now create a contingency table of the two variables Species and size with the `table()` function:

```{r, results=FALSE}
table(dat$Species, dat$size)
# or with the xtabs() function:
xtabs(~ dat$Species + dat$size)
```

- The contingency table gives the number of cases in each subgroup. 

---
- Note that Species are in rows and size in column because we specified Species and then size in table(). 

- Change the order if you want to switch the two variables.

- Instead of having the frequencies you can also have the relative frequencies (i.e., proportions) in each subgroup by adding the `table()` function inside the `prop.table()` function:
```{r, results=FALSE}
prop.table(table(dat$Species, dat$size))
```

- Note that you can also compute the percentages by row or by column by adding a second argument to the `prop.table()` function: 1 for row, or 2 for column:

```{r, eval=FALSE}
# percentages by row:
round(prop.table(table(dat$Species, dat$size), 1), 2) 
round(prop.table(table(dat$Species, dat$size), 2), 2) 
```

---

## Advanced descriptive statistics

- There are also, many more functions and packages to perform more advanced descriptive statistics in R. 

#### **summarytools** package

- One package for descriptive statistics I often use for my projects in R is the **summarytools()** package. The package is centered around 4 functions:

1.	`freq()` for frequencies tables
2.	`ctable()` for cross-tabulations
3.	`descr()` for descriptive statistics
4.	`dfSummary()` for dataframe summaries

- A combination of these 4 functions is usually more than enough for most descriptive analyses. 

---

#### Frequency tables with freq()

- The `freq()` function produces frequency tables with frequencies, proportions, as well as missing data information.

```{r, results=FALSE, message=FALSE, warning=FALSE}
library(summarytools)
freq(dat$Species)
```

- If you do not need information about missing values, add the `report.nas = FALSE` argument:

```{r, results=FALSE}
freq(dat$Species,
  report.nas = FALSE) # remove NA information
```

---

And for a minimalist output with only counts and proportions:
```{r, results=FALSE}
freq(dat$Species,
  report.nas = FALSE, # remove NA information
  totals = FALSE, # remove totals
  cumul = FALSE, # remove cumuls
  headings = FALSE) # remove headings
```

#### Cross-tabulations with ctable()

- The `ctable()` function produces cross-tabulations for pairs of categorical variables. Using the two categorical variables in our dataset:
```{r, results=FALSE}
ctable(x = dat$Species, y = dat$size)
```

---

- Row proportions are shown by default. To display column or total proportions, add the prop = "c" or prop = "t" arguments, respectively:
```{r, results=FALSE}
ctable( x = dat$Species,  y = dat$size,
  prop = "t" ) # total proportions
```

- To remove proportions altogether, add the argument prop = "n". 

- Furthermore, to display only the bare minimum, add the totals = FALSE and headings = FALSE arguments:
```{r, results=FALSE}
ctable(x = dat$Species, y = dat$size,
  prop = "n", # remove proportions
  totals = FALSE, # remove totals
  headings = FALSE) # remove headings
```

- This is equivalent than `table(dat$Species, dat$size)` and `xtabs(~ dat$Species + dat$size)` performed in the section on contingency tables.

---

- To display results of the Chi-square test of independence, add the chisq = TRUE argument:3
```{r, results=FALSE}
ctable(x = dat$Species, y = dat$size,
  chisq = TRUE, # display results of Chi-square test of independence
  headings = FALSE) # remove headings
```

- The p-value is close to 0 so we reject the null hypothesis of independence between the two variables. 
- This indicates that species and size are dependent and that there is a significant relationship between the two variables.

- It is also possible to create a contingency table for each level of a third categorical variable thanks to the combination of the `stby()` and `ctable()` functions. 

---

- There are only 2 categorical variables in our dataset, so let’s use the tabacco dataset which has 4 categorical variables (i.e., gender, age group, smoker, diseased). 

- For this example, we would like to create a contingency table of the variables smoker and diseased, and this for each gender:

```{r, results=FALSE}
stby(list(  x = tobacco$smoker, # smoker and diseased
  y = tobacco$diseased),
INDICES = tobacco$gender, # for each gender
FUN = ctable # ctable for cross-tabulation
)
```

---

#### Descriptive statistics with `descr()`

- The `descr()` function produces descriptive statistics with common central tendency and measures of dispersion. 

- A major advantage of this function is that it accepts single vectors as well as data frames. 

- If a data frame is provided, all non-numerical columns are ignored so you do not have to remove them yourself before running the function.

- The `descr()` function allows to display:
   - only a selection of some descriptive statistics, use `stats = c("mean", "sd")` argument.
   -	the min, 1st quartile, median, 3rd quartile and max, use `stats = "fivenum"`
  -	the most common descriptive statistics (mean, sd, min, median, max, valid observations), with `stats = "common"`:
  
---

```{r, results=FALSE}
descr(dat,headings = FALSE, # remove headings
  stats = "common" # most common descriptive statistics
)
```

Tip: if you have a large number of variables, add the` transpose = TRUE` argument for a better display.

- In order to compute these descriptive statistics by group (e.g., Species in our dataset), use the `descr()` function in combination with the `stby()` function:

```{r, results=FALSE}
stby( data = dat, INDICES = dat$Species, # by Species
  FUN = descr, # descriptive statistics
  stats = "common") # most common descr. stats
```

---

#### Data frame summaries with `dfSummary()`

- The `dfSummary()` function generates a summary table with statistics, frequencies and graphs for all variables in a dataset.

- The information shown depends on the type of the variables and also varies according to the number of distinct values.

```{r, results=FALSE}
dfSummary(dat)
```

#### `describeBy()` from the **psych()** package

- The `describeBy()` function from the `psych()` package allows to report several summary statistics 
  - (i.e., total valid cases, mean, sd, min, max, range, skewness and kurtosis) by a grouping variable.

---

```{r, results=FALSE, message=FALSE}
library(psych)
describeBy( dat,dat$Species) # grouping variable

```

#### `aggregate()` function

- The `aggregate()` function allows to split the data into subsets and then to compute summary statistics for each. 

- For instance, if we want to compute the mean for the variables Sepal.Length and Sepal.Width by Species and Size:
```{r, results=FALSE}
aggregate(cbind(Sepal.Length, Sepal.Width) ~ Species + size,
  data = dat, mean)
```

---

#### summaryBy() from **doBy()** package

- An alternative is the `summaryBy()` function from the `doBy()` package:

```{r, results=FALSE, message=FALSE}
# summary statistics by group
library(doBy)
summaryBy(Sepal.Length + Sepal.Width ~ Species,
  data = dat,  FUN = summary)
```

- If you are interested in some specific descriptive statistics, you can easily specify them via the FUN argument:

```{r, results=FALSE}
summaryBy(Sepal.Length + Sepal.Width ~ Species,
  data = dat,  FUN = c(mean, var))
```

---

#### group_by() and summarise() from {dplyr}

- Another alternative is with the `summarise()` and `group_by()` functions from the 

#### dplyr() package:

```{r, results=FALSE}
library(dplyr)
group_by(dat, Species) %>%
  summarise(mean = mean(Sepal.Length, na.rm = TRUE),
    sd = sd(Sepal.Length, na.rm = TRUE))
```


# Graphical Procedure 

## OVERVIEW

- Data visualization is part art and part science. 

- A data visualization first and foremost has to accurately convey the data. 

- It must not mislead or distort. 

- A data visualization should be aesthetically pleasing. 

- Good visual presentations tend to enhance the message of the visualization. 

- What are the key principles, methods, and concepts required to visualize data for publications, reports, or presentations? 

---

## Graphics in base R

- Class sensitive commands

- Everything can be adjusted through parameters.

- But sometimes hard to find the right ones.

- Require lots of trial-error, no easy patterns.

- Memorizing lots of plotting commands and options.

- Hard to manage, save, update, reproduce.

---

### Why ggplot2?

- A grammar of graphics is a grammar used to describe and create a wide range of statistical graphics.

- The promise of a **grammar for graphics**.

- Easy to manage, save, etc.

- Graphs are composed of layers.

- Easy to add stuff to existing graphs.

- ggplot2 graphics take less work to make beautiful and eye-catching graphics.

- Enables creation of reproducible visualization patterns.

- Publication quality & beyond

---

## Components of the layered grammar

- Layer
  - Data
  - Mapping
  - Statistical transformation (stat)
  - Geometric object (geom)
  - Position adjustment (position)
- Scale
- Coordinate system (coord)
- Faceting (facet)

---

### Layers

- Layers are used to create the objects on a plot.

- A layer is composed of four parts:

1. data and aesthetic mapping

2. geometric object (geom)

3. statistical transformation (stat)

4. position adjustment

- Layers are typically related to one another and share many common features. 

- example: a scatterplot overlayed with a smoothed regression line.

---

###  data and aesthetic mapping

#### Data

- Data defines the source of the information to be visualized.
- Must be a data.frame
- Gets pulled into the ggplot() object


#### Aesthetics (a.k.a. mapping)

- Defines how the variables are applied to the plot. 

- They are properties of graphical elements, such as: 
   - x, y coordinates of points
   - line type, size, shape, 
   - colour, fill colour of points, lines, arrows, bars, boxes, and so forth.
   - alpha level of transparency.

---

### 1. data and aesthetic mapping

- `ggplot(data = DATA, mapping = aes(MAPPINGS))+ GEOM_FUNCTION()`

```{r, eval=FALSE}
ggplot(dataset, aes(x=xvar, y=yvar)) + geom_function()
```

- **geom_function()** to produce shapes for the graph like: `geom_point()`, `geom_boxplot()`, ... 
- Inside **aes()**, we can map more variables like: color, size and shape of plotted objects.
.pull-left[
```{r,  fig.width=4.5, fig.height=2.4}
ggplot(mtcars,aes(x=mpg, y=hp))#<<
```
]
--
.pull-right[
```{r, fig.width=4.5, fig.height=2.4}
ggplot(mtcars, aes(x=mpg,y =hp)) +
  geom_point()
```
]

---


### Aesthetics: `aes()`

- `ggplot(data = DATA, mapping = aes(MAPPINGS))+ GEOM_FUNCTION()`
- Let us look what do `color`, `size `, `alpha` and `shape` in ggplot?


```{r, echo=FALSE, warning=FALSE}
diabetes <- read.csv("https://raw.githubusercontent.com/Yebelay/Statistical-computing/main/diabetes.csv")
```
.pull-left[
```{r, fig.height=4, fig.width=6, warning=FALSE, fig.align='center'}
ggplot(data = diabetes,mapping 
   = aes(x = weight, y = hip, 
         `color`=gender))+geom_point()
```
]
.pull-right[
```{r, fig.height=4, fig.width=6, warning=FALSE, fig.align='center'}
ggplot(data = diabetes,mapping = 
aes(x=weight,y=hip,`size`=gender))+
  geom_point()
```
]

---
.pull-left[
```{r, fig.height=4, fig.width=6, warning=FALSE, fig.align='center'}
ggplot(data = diabetes,mapping = 
aes(x=weight,y=hip,`alpha`=gender))+#<<
  geom_point()
```
]
.pull-right[
```{r, fig.height=4, fig.width=6, warning=FALSE, fig.align='center'}
ggplot(data = diabetes, mapping =
aes(x=weight,y=hip,`shape`=gender))+ #<<
  geom_point()
```
]

---

### geoms
**What shape does the data take?**

- `geom_point()`, `geom_line()`, `geom_violin()`

.pull-left[
```{r, fig.height=4, fig.width=6, warning=FALSE, fig.align='center'}
ggplot(diabetes,aes(gender,chol))+ 
  geom_boxplot() #<<
```
]
.pull-right[
```{r, fig.height=4, fig.width=6,message=FALSE, warning=FALSE, fig.align='center'}
ggplot(diabetes, aes(x = glyhb)) + 
  geom_histogram() #<<
```
]

---

.pull-left[
```{r, fig.height=4, fig.width=6,message=FALSE, warning=FALSE, fig.align='center'}
ggplot(diabetes, aes(x = glyhb)) + 
  geom_density() #<<
```
]
.pull-right[
```{r, fig.height=4, fig.width=6,message=FALSE, warning=FALSE, fig.align='center'}
diabetes %>% 
  ggplot(aes(x = frame)) + #<<
    geom_bar() #<<
```
]

---
.pull-left[
```{r, fig.height=4, fig.width=6,message=FALSE, warning=FALSE, fig.align='center'}
diabetes %>% 
  drop_na() %>% 
  ggplot(aes(x=frame, 
color=gender))+geom_bar()
```
]

.pull-right[
```{r, fig.height=4, fig.width=6,message=FALSE, warning=FALSE, fig.align='center'}
diabetes %>% 
  drop_na() %>% 
  ggplot(aes(x = frame, 
  fill=gender))+geom_bar()
```
]

---

## Positions

- `geom_bar(position = "<POSITION>")`

- When we have aesthetics mapped, how are they positioned?

- bar: dodge, fill, stacked (default)

- point: jitter

---

.pull-left[
```{r}
diabetes %>% 
  drop_na() %>% 
  ggplot(aes(x =frame, 
  fill=gender))+ 
geom_bar(position ="stack") #<<
```
]

.pull-right[
```{r}
diabetes %>% 
  drop_na() %>% 
  ggplot(aes(x = frame,
fill = gender)) + 
  geom_bar(position = "dodge") #<<
```
]

---

## Live Code Part 1

1. Predict what this code will do. Then run it. 
2. Add a `linetype` aesthetic for `gender`. Run it again.
3. Set the color of `geom_smooth()` to "black"
4. Add `se = FALSE` to the `geom_smooth()`
5. It's hard to see the lines well now. How about setting `alpha = .2` in `geom_point()`?
6. Jitter the points. You can either change the geom or change the `position` argument.
7. Add another layer, `theme_bw()`. Remember to use `+`.

```{r, eval=FALSE}
ggplot(diabetes, aes(weight, hip)) + 
  geom_point() +
  geom_smooth()
```

---

.pull-left[
```{r, fig.height=4, fig.width=6,message=FALSE, warning=FALSE, fig.align='center'}
ggplot(diabetes, aes(weight, hip)) + 
  geom_point() +
  geom_smooth()
```
]

.pull-right[
```{r, fig.height=4, fig.width=6,message=FALSE, warning=FALSE, fig.align='center'}
ggplot(diabetes, aes(weight, hip)) + 
  geom_point() +
  geom_smooth(aes(linetype = gender))
```
]

---

.pull-left[
```{r, fig.height=4, fig.width=6,message=FALSE, warning=FALSE, fig.align='center'}
ggplot(diabetes, aes(weight, hip)) + 
  geom_point() +geom_smooth(
aes(linetype = gender), 
col = "black")
```
]

.pull-right[
```{r, fig.height=4, fig.width=6,message=FALSE, warning=FALSE, fig.align='center'}
ggplot(diabetes, aes(weight, hip)) + 
  geom_point()+geom_smooth(
aes(linetype = gender), 
col = "black", se = FALSE)
```
]

---
.pull-left[
```{r, fig.height=4, fig.width=6,message=FALSE, warning=FALSE, fig.align='center'}
ggplot(diabetes, aes(weight, hip)) + 
  geom_point(alpha = .2) +
  geom_smooth(
  aes(linetype = gender), 
  col = "black", se = FALSE)
```
]
.pull-right[
```{r, fig.height=4, fig.width=6,message=FALSE, warning=FALSE, fig.align='center'}
ggplot(diabetes, aes(weight, hip)) + 
  geom_jitter(alpha = .2) +
  geom_smooth(aes(
    linetype = gender), 
    col = "black", se = FALSE)+ theme_bw()
```
]

---

### Facets

- Easy peazy panels
- `facet_grid()`
- `facet_wrap()`
- `x ~ y` or  ` ~ y` or  `x ~ .`

#### facet grid by `gender` and `location`

.pull-left[
```{r, eval=FALSE}
ggplot(diabetes, aes(weight, hip)) + 
  geom_point() + 
  geom_smooth() + 
  facet_grid(
    gender ~ location) #<<
```

]
.pull-right[
```{r, fig.height=3.5, fig.width=6,message=FALSE, warning=FALSE, fig.align='center', echo=FALSE}
ggplot(diabetes, aes(weight, hip)) + 
  geom_point() + geom_smooth() + 
  facet_grid(gender ~ location) #<<
```
]

---

### Themes

- Non-data ink (text, background, etc)

- Prespecified themes: `theme_gray()` (default), `theme_minimal()`, `theme_light()`, etc.
- `theme()`

.pull-left[
```{r, eval=FALSE, warning=FALSE, fig.align='center'}
mtcars %>% 
ggplot(aes(hp,mpg,
   col=factor(cyl)))+ 
geom_point(size=3) + 
theme(axis.text=
element_text(size = 8),  
legend.text=element_text(
  size=8,face="bold"),  
  legend.direction = "vertical")  
```
]

.pull-right[
```{r, fig.height=3.5, fig.width=6,message=FALSE, warning=FALSE, fig.align='center', echo=FALSE}
mtcars %>% 
ggplot(aes(hp, mpg, col = factor(cyl))) + 
  geom_point(size = 3) + 
  theme(axis.text = element_text(size = 8),  
        legend.text = element_text(size = 8, face = "bold"),  
        legend.direction = "vertical")  
```
]

---


## Labels, titles, and legends

### Add a title: 

- `ggtitle()`, `labs(title = "My Awesome Plot")`

### Change a label: 

-  `xlab()`, `ylab()`, `labs(x = "X!", y = "Y!!")`

---

```{r, fig.height=4, fig.width=7,message=FALSE, warning=FALSE, fig.align='center'}
diabetes_plot <- ggplot(diabetes, aes(weight, hip, linetype = gender)) + 
  geom_smooth(color = "black", se = FALSE) + 
  theme_bw(base_size = 12) +
  labs(x = "Weight (lbs)", y = "Hip (inches)") +
  ggtitle("Hip and Weight by Sex") + 
  scale_linetype(name = "Sex")
diabetes_plot
```
